Optimal (Multiway) Spatial Joins

### 2024-05-23 msj.tex
yufei

\begin{abstract}
    In a {\em spatial join}, we are given a constant number $k \geq 2$ of sets containing axis-parallel rectangles in a 2D space, denoted as $R_1, R_2, ..., R_k$. The objective is to report all $k$-tuples $(r_1, r_2, ..., r_k) \in R_1 \times R_2 \times ... \times R_k$ where the rectangles $r_1, r_2, ..., r_k$ have a non-empty intersection, i.e., $r_1 \cap r_2 \cap ... \cap r_k \neq \emptyset$. This problem holds significant importance in spatial databases and has been extensively studied for over two decades. We explain how to settle the problem in $O(n \log n + \out)$ time using $O(n + \out)$ space --- regardless of the constant $k$ --- where $n = \sum_{i=1}^k |R_i|$ and $\out$ is the result size (i.e., the total number of $k$-tuples reported). The time complexity is asymptotically optimal in the class of comparison-based algorithms, to which our solution belongs. Our result significantly improves the state of the art, which is an algorithm with running time $O(n \log^{2k} n + \out)$.
\end{abstract}

@@

From the perspective of theory, SJs are best understood when $k = 2$, i.e., the {\em pairwise} scenario, where it is folklore that the problem can be solved in $O(n \log n + \out)$ time using $O(n)$ space (e.g., by planesweep \cite{bcko08}). However, the problem becomes significantly more challenging for $k \ge 3$, i.e., the {\em multiway} scenario. All the solutions developed  before 2022 (see \cite{gcn+13,mp98,mp01,pmt99} and the references therein) suffer from a worst-case time complexity of $O(n^k)$, offering essentially no improvement over the naive method that enumerates the entire cartesian product $R_1 \times R_2 \times ... \times R_k$.

\begin{table*} 
    \begin{tabular}{c|c|c|c|c} 
        $\bm{k}$ & {\bf method} & {\bf runtime} & {\bf space} & {\bf remark} \\
        \hline\hline 
        2 & folklore & $O(n \log n + \out)$ & $O(n)$ & comparison-based optimal \\ 
        \hline
        $\ge 3$ & before 2022 & $O(n^k)$ & - & \\
        $\ge 3$ & \cite{ty22} & $O((n + \out) \cdot \polylog n)$ & $O(n \polylog n + \out)$ & \\
        $\ge 3$ & \cite{kcko22} & $O(n \log^{2k} n + \out)$ & $O(n \log n + \out)$ & \\
        \hline
        $\ge 3$ & ours & $O(n \log n + \out)$ & $O(n + \out)$ & comparison-based optimal
    \end{tabular}

    \vspace{3mm}
    \caption{Comparison of results on the $k$-SJ problem for a constant $k$}
    \label{tab:results-com}
    \figcapdown
\end{table*}

@@

Year 2022 witnessed two independent works \cite{ty22,kcko22} that, although not tackling $k$-SJ directly, imply provably fast $k$-SJ algorithms. Specifically, in \cite{ty22}, Tao and Yi studied several variants of ``interval intersection joins'' under updates. Most relevant to our context is the variant where the input includes, for each $i \in [k]$, a set $\I_i$ of 1D intervals in $\real$, and the join result comprises all $k$-tuples $(I_1,$ $I_2,$ $..., I_k) \in \I_1 \times \I_2 \times ... \times \I_k$ with $\bigcap_{i=1}^k I_i \neq \emptyset$. The objective is to design a data structure, which, given the insertion (resp., deletion) of an interval in one of the $k$ sets, can identify all the newly-appearing (resp., disappearing) $k$-tuples in the join result in $O((1+\Delta) \cdot \polylog n)$ time, where $n = \sum_{i=1}^k |\I_i|$ and $\Delta$ is the number of such $k$-tuples. Tao and Yi \cite{ty22} presented a structure of $O(n \polylog n)$ space achieving the purpose. Combining their structure with planesweep, one can obtain an algorithm for solving the $k$-SJ problem in $O((n + \out) \cdot \polylog n)$ time using $O(n \polylog n + \out)$ space.

@@

In \cite{kcko22}, Khamis et al.\ investigated a type of joins that extends the conventional equi-join in two ways. First, each attribute value in a relation is an interval (rather than a real value); second, each equality predicate in equi-join is replaced with a ``non-empty intersection'' predicate on the attributes involved. The $k$-SJ problem can be converted to solving a join defined next under the framework of \cite{kcko22}. For each $i \in [k]$, define $R_i$ as a relation over two attributes $X$ and $Y$. For each tuple $\bm{u} \in R_i$, its values $\bm{u}(X)$ and $\bm{u}(Y)$ on the two attributes are both intervals (effectively defining a rectangle). The objective is to output all $k$-tuples $(\bm{u}_1, \bm{u}_2, ..., \bm{u}_k) \in R_1 \times R_2 \times ... \times R_k$ satisfying $\bigcap_{i=1}^k \bm{u}_i(X) \ne \emptyset$ and $\bigcap_{i=1}^k \bm{u}_i(Y) \ne \emptyset$. It is clear that there is one-one correspondence between the result of this join and that of k-SJ. Khamis et al.\ \cite{kcko22} developed an algorithm that can process the join  in $O(n \log^{2k} n + \out)$ time. Their algorithm requires $O(n \log n + \out)$ space.

@@

@@

In this paper, we show that the $k$-SJ problem can be settled by a comparison-based algorithm that runs in $O(n \log n + \out)$ time and uses  $O(n + \out)$ space, regardless of the constant $k$. The time complexity is asymptotically optimal.

@@

We will also prove that the space of our algorithm is $O(k \cdot n + k \cdot \out)$. When $k = O(1)$, the time and space complexities become $O(n \log n$ $+ \out)$ and $O(n + \out)$, as promised.

@@

In Appendix~\ref{app:prob-a}, we explain how to solve the problem in $O(n \log n)$ time and $O(n)$ space where $n = |H| + |R|$.

@@

In Appendix~\ref{app:prob-b}, we explain how to solve the problem in $O(n \log n)$ time and $O(n)$ space where $n = |H| + |R|$.

@@

In Appendix~\ref{app:prob-c}, we explain how to solve the problem in $O(n \log n + \out)$ time and $O(n + \out)$ space where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

In Appendix~\ref{app:prob-d}, we explain how to solve the problem in $O(n \log n + \out)$ time and $O(n + \out)$ space where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $h$ crosses $r$.

@@

Our objective is to prove that the H-V $k$-SJ problem can be efficiently reduced to the $(k-1)$-SJ problem --- note: it is $(k-1)$-SJ here, rather than H-V $(k-1)$-SJ. For this purpose, we will assume the existence of an algorithm $\A$ that can solve $\kappa$-SJ problems for all $\kappa \in [2, k-1]$. Denote by
\myitems{
    \item $F_\kappa(n, \out)$ the worst-case runtime of $\A$ on any instance of $\kappa$-SJ that has input size $n$ and output size $\out$;
    \item $G_\kappa(n, \out)$ the worst-case space consumed by $\A$ on any instance mentioned above.
}
We consider that $F_\kappa(n, \out) \le F_{\kappa + 1}(n, \out)$ and $G_\kappa(n, \out) \le G_{\kappa + 1}(n, \out)$ for any $\kappa \in [2, k-2]$, that is, its overhead on $\kappa$-SJ should not be larger than that on $(\kappa+1)$-SJ.

@@

\begin{lemma} \label{lmm:hv}
    The H-V $k$-SJ problem can be solved in 
    \myeqn{
        O(k) \cdot \big( F_{k-1}(n, \out) + n\log n + k \cdot \out \big) \nn 
    }
    time using $O(n + \out) + G_{k-1}(n, \out)$ space.
\end{lemma}

@@

\noindent In Appendix~\ref{app:hv:type1:gen-S}, we prove that the set $S$ computed satisfies \eqref{eqn:hv:type1:S}, as desired.

>>>>

\begin{abstract}
    In a {\em spatial join}, we are given a constant number $k \geq 2$ of sets containing axis-parallel rectangles in a 2D space, denoted as $R_1, R_2, ..., R_k$. The objective is to report all $k$-tuples $(r_1, r_2, ..., r_k) \in R_1 \times R_2 \times ... \times R_k$ where the rectangles $r_1, r_2, ..., r_k$ have a non-empty intersection, i.e., $r_1 \cap r_2 \cap ... \cap r_k \neq \emptyset$. This problem holds significant importance in spatial databases and has been extensively studied for over two decades. We explain how to settle the problem in $O(n \log n + \out)$ time --- regardless of the constant $k$ --- where $n = \sum_{i=1}^k |R_i|$ and $\out$ is the result size (i.e., the total number of $k$-tuples reported). The time complexity is asymptotically optimal in the class of comparison-based algorithms, to which our solution belongs. Our result significantly improves the state of the art, which is an algorithm with running time $O(n \log^{2k} n + \out)$.
\end{abstract}

@@

From the perspective of theory, SJs are best understood when $k = 2$, i.e., the {\em pairwise} scenario, where it is folklore that the problem can be solved in $O(n \log n + \out)$ time (e.g., by planesweep \cite{bcko08}). However, the problem becomes significantly more challenging for $k \ge 3$, i.e., the {\em multiway} scenario. All the solutions developed  before 2022 (see \cite{gcn+13,mp98,mp01,pmt99} and the references therein) suffer from a worst-case time complexity of $O(n^k)$, offering essentially no improvement over the naive method that enumerates the entire cartesian product $R_1 \times R_2 \times ... \times R_k$.

@@

Year 2022 witnessed two independent works \cite{ty22,kcko22} that, although not tackling $k$-SJ directly, imply provably fast $k$-SJ algorithms. Specifically, in \cite{ty22}, Tao and Yi studied several variants of ``interval intersection joins'' under updates. Most relevant to our context is the variant where the input includes, for each $i \in [k]$, a set $\I_i$ of 1D intervals in $\real$, and the join result comprises all $k$-tuples $(I_1,$ $I_2,$ $..., I_k) \in \I_1 \times \I_2 \times ... \times \I_k$ with $\bigcap_{i=1}^k I_i \neq \emptyset$. The objective is to design a data structure, which, given the insertion (resp., deletion) of an interval in one of the $k$ sets, can identify all the newly-appearing (resp., disappearing) $k$-tuples in the join result in $O((1+\Delta) \cdot \polylog n)$ time, where $n = \sum_{i=1}^k |\I_i|$ and $\Delta$ is the number of such $k$-tuples. Tao and Yi \cite{ty22} presented a structure of $O(n \polylog n)$ space achieving the purpose. Combining their structure with planesweep, one can obtain an algorithm for solving the $k$-SJ problem in $O((n + \out) \cdot \polylog n)$ time.

@@

In \cite{kcko22}, Khamis et al.\ investigated a type of joins that extends the conventional equi-join in two ways. First, each attribute value in a relation is an interval (rather than a real value); second, each equality predicate in equi-join is replaced with a ``non-empty intersection'' predicate on the attributes involved. The $k$-SJ problem can be converted to solving a join defined next under the framework of \cite{kcko22}. For each $i \in [k]$, define $R_i$ as a relation over two attributes $X$ and $Y$. For each tuple $\bm{u} \in R_i$, its values $\bm{u}(X)$ and $\bm{u}(Y)$ on the two attributes are both intervals (effectively defining a rectangle). The objective is to output all $k$-tuples $(\bm{u}_1, \bm{u}_2, ..., \bm{u}_k) \in R_1 \times R_2 \times ... \times R_k$ satisfying $\bigcap_{i=1}^k \bm{u}_i(X) \ne \emptyset$ and $\bigcap_{i=1}^k \bm{u}_i(Y) \ne \emptyset$. It is clear that there is one-one correspondence between the result of this join and that of k-SJ. Khamis et al.\ \cite{kcko22} developed an algorithm that can process the join  in $O(n \log^{2k} n + \out)$ time.

@@

\begin{table} 
    \begin{tabular}{c|c|c|c} 
        $\bm{k}$ & {\bf method} & {\bf runtime} & {\bf remark} \\
        \hline\hline 
        2 & folklore & $O(n \log n + \out)$ & optimal \\ 
        \hline
        $\ge 3$ & before 2022 & $O(n^k)$ &  \\
        $\ge 3$ & \cite{ty22} & $O((n + \out) \cdot \polylog n)$ & \\
        $\ge 3$ & \cite{kcko22} & $O(n \log^{2k} n + \out)$ & \\
        \hline
        $\ge 3$ & ours & $O(n \log n + \out)$ & optimal
    \end{tabular}

    \vspace{3mm}
    \figcapup 
    \caption{Result comparison on $k$-SJ problem for a constant $k$}
    \label{tab:results-com}
    \figcapdown \vspace{-5mm}
\end{table}

@@

In this paper, we show that the $k$-SJ problem can be settled by a comparison-based algorithm that runs in $O(n \log n + \out)$ time, regardless of the constant $k$. The time complexity is asymptotically optimal.

@@

When $k = O(1)$, the time complexity becomes $O(n \log n$ $+ \out)$, as promised.

@@

In Appendix~\ref{app:prob-a}, we explain how to solve the problem in $O(n \log n)$ time where $n = |H| + |R|$.

@@

In Appendix~\ref{app:prob-b}, we explain how to solve the problem in $O(n \log n)$ time where $n = |H| + |R|$.

@@

In Appendix~\ref{app:prob-c}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

In Appendix~\ref{app:prob-d}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $h$ crosses $r$.

@@

Our objective is to prove that the H-V $k$-SJ problem can be efficiently reduced to the $(k-1)$-SJ problem --- note: it is $(k-1)$-SJ here, rather than H-V $(k-1)$-SJ. For this purpose, we will assume the existence of an algorithm $\A$ that can solve $\kappa$-SJ problems for all $\kappa \in [2, k-1]$. Denote by $F_\kappa(n, \out)$ the worst-case runtime of $\A$ on any instance of $\kappa$-SJ that has input size $n$ and output size $\out$. We consider that $F_\kappa(n, \out) \le F_{\kappa + 1}(n, \out)$, that is, its overhead on $\kappa$-SJ should not be larger than that on $(\kappa+1)$-SJ.

@@

\begin{lemma} \label{lmm:hv}
    The H-V $k$-SJ problem can be solved in time
    \myeqn{
        O(k) \cdot \big( F_{k-1}(n, \out) + n\log n + k \cdot \out \big). \nn 
    }
\end{lemma}

@@

\noindent In Appendix~\ref{app:hv:type1:gen-S}, we prove that the set $S$ computed satisfies \eqref{eqn:hv:type1:S}, as desired. The running time of the above algorithm is
\myeqn{
    O(1 + |\Pi_{1:k-2}(\J_1(R_1, ..., R_{k-2}, H, V))| + |S|) = O(1 + \out) \nn
}
and its space is bounded by $O(\out)$.
